Clinical prediction using similarity cohort--based models: A systematic scoping review

### 2025-03-09 docs/Localized-Modeling-BMCAI/localized-modeling-dai-revision.tex
write additional paragraphs + visualize performance dependencies

This approach harkens to the aspirational ``green button'' that would,
in response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient \citep{Longhurst2014}. Several hurdles
face the deployment of such an approach in practice, but its benefits
must first be demonstrated. Our motivations in this scoping review are
to describe the settings in and problems with which this approach has
been tasked, to attempt to organize them within a common framework, and
to evaluate the promise they show toward achieving this or other needs
in clinical informatics. Along the way, we attempt to reconcile
terminology, summarize motivations and evaluations, and propose valuable
follow-up work.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems \citep{Aamodt1994}. Early
implementations realized a general workflow described as ``the four
REs'', later the ``R4 cycle'' \citep{Aamodt1994, Begum2011}: Given a new
case (or problem), the system \emph{retrieves} one or more past cases
(solved problems) from a corpus, \emph{reuses} these to generate an
understanding of (solution to) the new case, \emph{revises} this
understanding (solution) to better fit the new case (sometimes called
``adaptation''), and \emph{retains} the new case and its eventual
understanding (solution) in the corpus to be retrieved and reused in
future. In contrast to rule-based systems, which are variable-based and
generally interpretable as a single rule applied to all new cases,
case-based systems provide case-specific interpretations in the form of
a number of more fully understood reference cases.

\citet{Kolodner1992} distinguished two styles of CBR:
\emph{problem-solving}, which is more procedural and used when
objectives are more clearly defined, and \emph{interpretative}, which
provides categorizations and justifications for possible solutions. A
similar dichotomy is commonly used to distinguish the \emph{performance}
\hl{criterion }of predictive models from the \emph{interpretability}
criterion\hl{.} In these terms, \hl{SCBM}s are highly procedural: The
step of fitting a predictive model to a retrieved cohort is an automated
adaptive strategy \citep{Begum2011}, and the parameters that govern
cohort retrieval can be tuned alongside model hyperparameters in a
machine learning (ML) workflow. Accordingly, \hl{SCBM}s have been
primarily used for and evaluated on their ability to predict classes or
outcomes. Nevertheless, as \hl{we find in this review}, they can be used
to draw inferences about the importance of different risk factors to
specific individuals.

@@

We refer the interested reader to several previous reviews of CBR in
medicine \citep{Gierl1998, Begum2011, Choudhury2016}, of measures of
patient similarity \citep{Dai2020}, and of uses of patient similarity in
predictive models \citep{Welch2013, Sharafoddini2017, Parimbelli2018}.
While the reviews of CBR focus on applications using health data, the
patient similarity reviews encompass many additional types of data
(various molecular -omics, genetic tests, medical images, laboratory
tests, patient preferences, patient-reported outcomes, tracking devices,
social media) and survey a much broader scope of models (exploratory
analysis via dimension reduction and cluster analysis; risk evaluation
and outcome prediction; \hl{CDSS} and software tools). For the present
review, we are interested in how similarity matching on patient-level
health data is used to construct local cohorts for predictive or
inferential
modeling.\hl{ While the similarity measure used in the retrieval step need not be changed as the corpus grows, proposed measures have been diverse, contested, and rarely systematically compared.}

@@

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While these studies trace back to the late 1990s, they
have become \hl{slightly }more common, which suggests that this is an
active, though not rapidly expanding, approach. We note that CBR in
health and medicine originated as early as 1990, and \hl{SCBM} emerged
soon after CBR had established itself; the idea has been ``in the air''
for as long as CBR has been in use.\hl{ Based on the motivations and methodologies observed in our sample (discussed in later sections), we speculate that the higher frequency of studies since 2013 is attributable in large part to the availability of vaster clinical databases with improved pre-processing tools, following the rapid uptake of EHRs; and of more efficient and powerful tools for retrieval (similarity-matching), the traditional bottleneck of CBR. Another possibility is an increased interest in interpretable machine learning }\citep{Rudin2022}\hl{. Each of these factors should apply with no less force to other CDSS and clinical applications of ML, and the rise we see may simply be commensurate with an overall increase in the volume of articles in these fields.}

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1154}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1538}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1209}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1978}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1538}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1044}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1538}}@{}}

@@

560; 760 & 35; 14; 31; 8 \\

@@

Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 1,000; 447 & 6;
6 \\

@@

Imaging\hspace{6em} & Cross-sectional\hspace{6em} & 810; 302 & 8 \\

@@

@@

These programs used varying terminology for common concepts, and no
common term is in use for what we here refer to as
\hl{similarity cohort--based modeling}; authors described their
approaches as ``targeted prognosis'' \citep{Mariuzzi1997},
``transductive inference'' \citep{Song2006}, ``personalized decision
support'' \citep{Lee2015}, ``personalized (predictive) models''
\citetext{\citealp[in contrast to ``local
models'']{Liang2015}; \citealp{Ng2015}; \citealp{Wang2019}; \citealp{Ma2020}; \citealp{Liu2022}; \citealp{Doborjeh2022}},
and ``precision cohort'' analysis \citep{Wang2019, Tang2021, Ng2021}. We
find uses of ``targeted'', ``personalized'', and ``precision'' generic
and imprecise, while transductive inference is an established term for a
broader set of methods in ML.
\hl{While ``local'' naturally contrasts with ``global'', the overloaded term ``localized models'' stands to cause more confusion than it resolves, so we use ``similarity cohort--based models'' to refer to this specific approach.}

@@

%\textbf{TODO: Place vertical separators between major columns.}

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1088}} |

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1658}} |

@@

  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0984}} |

@@

& & & & \hspace{6em} & \hl{kNN} & 88.29\%\hspace{6em} \\

@@

& & & & \hspace{6em} & \hl{kNN} & 82.22\%\hspace{6em} \\

@@

& & Breast & Statistical CBR & 96.96\%\hspace{6em} & \hl{kNN} &

@@

& & & & \hspace{6em} & \hl{kNN} & 65.39\%\hspace{6em} \\

@@

& & & & \hspace{6em} & \hl{kNN} & 55.59\%\hspace{6em} \\

@@

& & Crohn's disease risk & TWNFI & 82.45\%\hspace{6em} & &

@@

\citet{Malykh2018} & Accuracy & J13/pneumonia & Case-based \hl{ANN} &
81.6\%\hspace{6em} & & \hspace{6em} \\
& & K80.1/gallbladder & Case-based \hl{ANN} & 76.7\%\hspace{6em} & &
& & H25.1/age-related cataract & Case-based \hl{ANN} & 94.9\%\hspace{6em} & &
& & H26.2/complicated cataract & Case-based \hl{ANN} & 91.4\%\hspace{6em} & &
& & 167.4/encephalopathy & Case-based \hl{ANN} & 72.4\%\hspace{6em} & &
& & 167.9/CBD & Case-based \hl{ANN} & 75.4\%\hspace{6em} & & \hspace{6em} \\
& & N20.1/ureter & Case-based \hl{ANN} & 58.7\%\hspace{6em} & &

>>>>

\hl{Interpretability is often coupled and contrasted with \emph{explainability}.
We follow }\citet{Lisboa2023}\hl{ in distinguishing these properties in terms of \emph{ante-hoc} versus \emph{post-hoc} model design and analysis choices, though we note that this usage is not universal }\citep{Molnar2023}\hl{. Work on explainable ML (or AI) is driven by many of the same needs as interpretable ML, including SCBM, in particular extracting knowledge from routinely collected data and balancing improvements in efficiency and accuracy with user or client comprehension and trust }\citep{Lisboa2023}\hl{.
However, we find compelling the case that interpretable models are preferable to explainable models when their performance is commensurate, especially for high-stakes decisions, due to the often shallow and fragile insights offered by explainers in comparison to models with transparent and comprehensible structure }\citep{Rudin2022}\hl{.
This review is driven by our interest specifically in interpretable ML, and by our suspicion that cohort retrieval has more potential for exploratory and predictive modeling of EHR and other routinely collected health data than is yet being exploited.}

\hl{SCBM} harkens to the aspirational ``green button'' that would, in
response to a query, automatically retrieve patient data from an
institution's records with which to conduct on-demand retrospective
studies for an individual patient \citep{Longhurst2014}. Several hurdles
face the deployment of such an approach in practice, but its benefits
must first be demonstrated. Our motivations in this scoping review are
to describe the settings in and problems with which this approach has
been tasked, to attempt to organize them within a common framework, and
to evaluate the promise they show toward achieving this or other needs
in clinical informatics. Along the way, we attempt to reconcile
terminology, summarize motivations and evaluations, and propose valuable
follow-up work.

@@

Clinical CBR emerged among rule-based approaches and other AI tools in
the development of expert systems \citep{Aamodt1994}. Early
implementations realized a general workflow described as ``the four
REs'', later the ``R4 cycle'' \citep{Aamodt1994, Begum2011}: Given a new
case (or problem), the system \emph{retrieves} one or more past cases
(solved problems) from a corpus, \emph{reuses} these to generate an
understanding of (solution to) the new case, \emph{revises}
\hl{or \emph{adapts}}this understanding (solution) to better fit the new
\hl{case,} and \emph{retains} the new case and its eventual
understanding (solution) in the corpus to be retrieved and reused in
future. In contrast to rule-based systems, which are variable-based and
generally interpretable as a single rule applied to all new cases,
case-based systems provide case-specific interpretations in the form of
a number of more fully understood reference cases.

\citet{Kolodner1992} distinguished two styles of CBR:
\emph{problem-solving}, which is more procedural and used when
objectives are more clearly defined, and \emph{interpretative}, which
provides categorizations and justifications for possible solutions. A
similar dichotomy is commonly used to distinguish the \emph{performance}
\hl{criterion }of predictive models from the \emph{interpretability}
criterion\hl{.}
\citet{Rudin2022}\hl{ distinguish \emph{prototype-based} CBR, based on the selection or creation of prototype cases for more efficient matching, from classical \emph{nearest neighbor--based} CBR.}
In these terms, \hl{SCBM}s are \hl{nearest neighbor--based and }highly
procedural: The step of fitting a predictive model to a retrieved cohort
is an automated adaptive strategy \citep{Begum2011}, and the parameters
that govern cohort retrieval can be tuned alongside model
hyperparameters in a machine learning (ML) workflow. Accordingly,
\hl{SCBM}s have been primarily used for and evaluated on their ability
to predict classes or outcomes. Nevertheless, as
\hl{we find in this review}, they can be used to draw inferences about
the importance of different risk factors to specific individuals.

@@

We refer the interested reader to several previous reviews of CBR in
medicine
\citep{Gierl1998, Begum2011, Choudhury2016}\hl{, of CDSS more broadly }\citep{Wasylewicz2019}\hl{ and of AI-driven systems }\citep{Elhaddad2024},
\hl{of patient similarity measures}
\citep{Dai2020}\hl{, of kNN algorithms }\citep{Halder2024},
\hl{of theory and uses of CBR for} predictive models
\citep{Welch2013, Sharafoddini2017, Parimbelli2018, Badra2023}\hl{, and of recent advances in CBR with emphasis on interpretability and explainability }\citep{Schoenborn2021, Rudin2022}.
While the reviews of CBR \hl{and CDSS }focus on applications using
health data, the \hl{others consider }additional types of data (various
molecular -omics, genetic tests, medical images, laboratory tests,
patient preferences, patient-reported outcomes, tracking devices, social
media) and survey a much broader scope of models (exploratory analysis
via dimension reduction and cluster analysis; risk evaluation and
outcome prediction\hl{; }software tools).

For the present review, we are interested in how similarity matching on
patient-level health data is used to construct local cohorts for
predictive or inferential
modeling.\hl{ While the similarity measure used in the retrieval step need not be changed as the corpus grows, proposed measures have been diverse, contested, and rarely systematically compared. Moreover, we found no reviews or surveys that focus specifically on the kinds of models we call SCBMs or include detailed comparisons between them.}

@@

Another interesting pattern lay in the years of publication (Figure
\ref{fig:year}). While these studies trace back to the late 1990s, they
have become \hl{slightly }more common, which suggests that this is an
active, though not rapidly expanding, approach. We note that CBR in
health and medicine originated as early as 1990, and \hl{SCBM} emerged
soon after CBR had established itself; the idea has been ``in the air''
for as long as CBR has been in use.
\hl{ Based on the motivations and methodologies observed in our sample (discussed in later sections), we speculate that the higher frequency of studies since 2013 is attributable in large part to the availability of vaster clinical databases with improved pre-processing tools, following the rapid uptake of EHRs; and of more efficient and powerful tools for retrieval (similarity-matching), the traditional bottleneck of CBR. Another possibility is an increased interest in interpretable machine learning }\citep{Rudin2022}\hl{. Each of these factors should apply with no less force to other CDSS and clinical applications of ML, and the rise we see may simply be commensurate with an overall increase in the volume of articles in these fields.}

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1123}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1497}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1176}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1925}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1497}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1283}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1497}}@{}}

@@

560; 760; 340 & 35; 14; 31; 8; 7 \\

@@

Clinical\hspace{6em} & Cross-sectional\hspace{6em} & 1,000 & 6 \\

@@

Imaging\hspace{6em} & Cross-sectional\hspace{6em} & 302 & 8 \\

@@

Figure S1 \textbf{TODO}.

@@

These programs used varying terminology for common concepts, and no
common term is in use for what we here refer to as
\hl{similarity cohort--based modeling}; authors described their
approaches as ``targeted prognosis'' \citep{Mariuzzi1997},
``transductive inference'' \citep{Song2006}, ``personalized decision
support'' \citep{Lee2015}, ``personalized (predictive) models''
\citetext{\citealp[in contrast to ``local
models'']{Liang2015}; \citealp{Ng2015}; \citealp{Wang2019}; \citealp{Ma2020}; \citealp{Liu2022}; \citealp{Doborjeh2022}},
and ``precision cohort'' analysis \citep{Wang2019, Tang2021, Ng2021}.
\hl{(Interestingly, none of the articles described their methods as ``explainable'' ML.)}
We find uses of ``targeted'', ``personalized'', and ``precision''
generic and imprecise, while transductive inference is an established
term for a broader set of methods in ML.
\hl{While \emph{local} naturally contrasts with \emph{global}, the overloaded term \emph{localized models} stands to cause more confusion than it resolves, so we use \emph{similarity cohort--based models} to refer to this specific approach.}

@@

\textbf{TODO: Place vertical separators (\texttt{\textbar{}}) between
major columns.}

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1088}}

@@

  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1658}}

@@

  >{\raggedleft\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0984}}

@@

& & & & \hspace{6em} & kNN & 88.29\%\hspace{6em} \\

@@

& & & & \hspace{6em} & kNN & 82.22\%\hspace{6em} \\

@@

& & Breast & Statistical CBR & 96.96\%\hspace{6em} & kNN &

@@

& & & & \hspace{6em} & kNN & 65.39\%\hspace{6em} \\

@@

& & & & \hspace{6em} & kNN & 55.59\%\hspace{6em} \\

@@

@@

\citet{Malykh2018} & Accuracy & J13/pneumonia & & \hspace{6em} &
Case-based ANN & 81.6\%\hspace{6em} \\
& & K80.1/gallbladder & & \hspace{6em} & Case-based ANN &
76.7\%\hspace{6em} \\
& & H25.1/age-related cataract & & \hspace{6em} & Case-based ANN &
94.9\%\hspace{6em} \\
& & H26.2/complicated cataract & & \hspace{6em} & Case-based ANN &
& & 167.4/encephalopathy & & \hspace{6em} & Case-based ANN &
72.4\%\hspace{6em} \\
& & 167.9/CBD & & \hspace{6em} & Case-based ANN & 75.4\%\hspace{6em} \\
& & N20.1/ureter & & \hspace{6em} & Case-based ANN &
58.7\%\hspace{6em} \\
